---
title: Codebook for Getting and Cleaning Data Course final project
author: Antonin Paquet
date: 2021-07-14
---

## Project Description
This is the final project for the Getting and Cleaning Data Course. It aimed at tiding a subset of the Human Activity Recognition Using Smartphones Dataset Version 1.0 (UCI HAR Dataset) and producing a second tidy data set with the average of each variable for each activity and each subject.

## Information on the raw data
The raw data used for this project was the UCI HAR Dataset. The raw data came from three axial signals of acceleremeters and gyroscopes that the participant were wearing while performing a series of activities (walking, walking upstairs, walking downstairs, sitting, standing, laying). Following is the explanation of the variables as quoted directly from the features_info.txt accompaying the UCI HAR Dataset :  

> The features selected for this database come from the accelerometer and gyroscope 3-axial raw signals tAcc-XYZ and tGyro-XYZ. These time domain signals (prefix 't' to denote time) were captured at a constant rate of 50 Hz. Then they were filtered using a median filter and a 3rd order low pass Butterworth filter with a corner frequency of 20 Hz to remove noise. Similarly, the acceleration signal was then separated into body and gravity acceleration signals (tBodyAcc-XYZ and tGravityAcc-XYZ) using another low pass Butterworth filter with a corner frequency of 0.3 Hz.
>
Subsequently, the body linear acceleration and angular velocity were derived in time to obtain Jerk signals (tBodyAccJerk-XYZ and tBodyGyroJerk-XYZ). Also the magnitude of these three-dimensional signals were calculated using the Euclidean norm (tBodyAccMag, tGravityAccMag, tBodyAccJerkMag, tBodyGyroMag, tBodyGyroJerkMag). 
>
Finally a Fast Fourier Transform (FFT) was applied to some of these signals producing fBodyAcc-XYZ, fBodyAccJerk-XYZ, fBodyGyro-XYZ, fBodyAccJerkMag, fBodyGyroMag, fBodyGyroJerkMag. (Note the 'f' to indicate frequency domain signals). 
>
These signals were used to estimate variables of the feature vector for each pattern:  
'-XYZ' is used to denote 3-axial signals in the X, Y and Z directions.
>
tBodyAcc-XYZ   
tGravityAcc-XYZ  
tBodyAccJerk-XYZ  
tBodyGyro-XYZ  
tBodyGyroJerk-XYZ  
tBodyAccMag  
tGravityAccMag  
tBodyAccJerkMag  
tBodyGyroMag  
tBodyGyroJerkMag  
fBodyAcc-XYZ  
fBodyAccJerk-XYZ  
fBodyGyro-XYZ  
fBodyAccMag  
fBodyAccJerkMag  
fBodyGyroMag  
fBodyGyroJerkMag  
>
The set of variables that were estimated from these signals are:
>
mean(): Mean value  
std(): Standard deviation  
mad(): Median absolute deviation   
max(): Largest value in array  
min(): Smallest value in array  
sma(): Signal magnitude area  
energy(): Energy measure. Sum of the squares divided by the number of values.   
iqr(): Interquartile range  
entropy(): Signal entropy  
arCoeff(): Autorregresion coefficients with Burg order equal to 4  
correlation(): correlation coefficient between two signals  
maxInds(): index of the frequency component with largest magnitude  
meanFreq(): Weighted average of the frequency components to obtain a mean frequency  
skewness(): skewness of the frequency domain signal  
kurtosis(): kurtosis of the frequency domain signal  
bandsEnergy(): Energy of a frequency interval within the 64 bins of the FFT of each window.  
angle(): Angle between to vectors.
>
Additional vectors obtained by averaging the signals in a signal window sample. These are used on the angle() variable:
>
gravityMean  
tBodyAccMean  
tBodyAccJerkMean  
tBodyGyroMean  
tBodyGyroJerkMean  

## Creating the tidy datafile

### Creating the tidy features_average file
1. Download the [UCI HAR Dataset](https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip) 
2. Run the run_analysis.R script from a folder containing the UCI HAR Dataset 
folder

### Processing and tidying of the data
The run_analysis.R script combines the various tables in the UCI HAR Dataset to 
add to the measured values table : descriptive variables names, a column with descriptive activity names and a column with subject id. A column for a **set** variable is also added, with values **train** or **test**, before combining the train and test data, in case that information needs to be retrieved later. The scripts also extracts the measurements on the mean and standard deviation for each measurement as was asked in the project instructions. (It should be noted that these project instructions were interpreted to mean to extract the values that calculated the mean and standard deviation as the last step, so only the features with "mean()" and "std()" at the end of the names were extracted.). Finaly, the script creates a second tidy data set, named *features_averages.txt*, with the averages for the different features, for each activities and each sujects. The detail of the script working can be found in the README.md file accomganying this data set. 

## Description of the variables in the tidy features_averages.txt file
```{r, echo = FALSE}
library(knitr)
averages <- read.table("features_averages.txt", header = TRUE)
kable(dim(averages), caption = "Dimensions of the data set", col.names = NULL)
kable(summary(averages)[, 1:6], caption = "Summary of the data set")
kable(names(averages), caption = "Variables present in the data set", col.names = NULL)
```

### Fixed variable 1
Name : subject_id  
Describes the subject id within a group of 30 volunteers who participated in the study  
 
 - Class of the variable :  `r class(averages[, 1])`  
 - Unique values/levels of the variable : `r unique(averages[, 1])` 
 - Unit of measurement : no units 
 
### Fixed variable 2
Name : activity  
Describes the activities that the subjects were performing during the measurements
 
 - Class of the variable :  `r class(averages[, 2])`  
 - Unique values/levels of the variable : `r unique(averages[, 2])` 
 - Unit of measurement : no units 

### Fixed variable 3
Name : set  
Describes the set that subjects were assigned to
 
 - Class of the variable :  `r class(averages[, 3])`  
 - Unique values/levels of the variable : `r unique(averages[, 3])` 
 - Unit of measurement : no units 

### Measured variables 1 to 66

These variables are named according to multiple levels of descriptor tokens that refer to the type of measurements and the various operations applied on the measurements. 

#### descriptors tokens used in the measured variable names
```{r, echo = FALSE}
library(knitr)
descriptors <- dplyr::tibble("Descriptor token used in raw data" = 
           c("t", "f", "Body", "Gravity", "Acc", "Gyro", "Jerk", "-X", "-Y",
             "-Z", "Mag", "-mean()", "-std"), 
       "New descriptor token used in tidy data" = 
           c("time", "frequency", "body", "gravity", "acc", "gyro", "jerk", "x", 
             "y", "z", "mag", "mean", "std"),
       "Description" = 
           c("Refers to time domain signal", "Refers to frequency domain signal",
             "Refers to the portion of the acceleration signal that is associated with the motion of the body",
             "Refers to the portion of the acceleration signal that is associated with the effect of the  earth’s gravity",
             "Refers to raw linear acceleration signal from accelerometer or signals derived from that raw signal",
             "Refers to raw angular velocity signal form gyroscope or signals derived from that raw signal",
             "Refers to jerk signal calculated by derivation of linear acceleration or angular velocity in time",
             "Refers to a signal axial component in the x direction",
             "Refers to a signal axial component in the y direction",
             "Refers to a vector signal axial component in the z direction",
             "Refers to the magnitude of a vector signal calculated with Euclidean norm", 
             "Refers to the mean value variable estimated from a signal",
             "Refers to the standard deviation variable estimated from a signal"))
kable(descriptors)
```

 - Class of the variables :  `r class(averages[, 4])` 
 - Unique values/levels of the variable : as many as lines
 - Unit of measurement : "The Features are normalized and bounded within [-1,1]" (ref : README.txt file accompanying UCI HAR Dataset)

### additional notes on measured variable names formating
All variable names were converted to lower caps underscore (snake) formatting. Special characters like "-" and "()" were removed. 



